\textbf{ Normalized Gaussian Processes }

### 2020-11-30 tex/ms.tex
better notation

In this note, we show that these two common practices are somewhat at odds
with each other. Specifically, if a physical process that generates
a dataset is distributed as a GP, the normalized process \emph{cannot} be
distributed as a GP. Provided certain conditions are met, the normalized
process can be well \emph{approximated} by a GP, albeit one with a different
covariance matrix $\tilde{\pmb{\Sigma}}$ that is not simply a scalar
multiple of the original covariance matrix. Moreover, if the original process
is described by a stationary kernel (i.e., one in which covariances are
independent of phase), the normalized process is not guaranteed to be.

@@

Let $\mathbf{x} = \left( x_0 \,\, \cdots \,\, x_{K-1} \right)^\top$
be a $K$-dimensional multivariate normal random variable distributed
according to a Gaussian Process with mean $\pmb{\mu}$
and covariance $\pmb{\Sigma}$:
%
\begin{align}
    \mathbf{x} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad.
\end{align}
%
For simplicity, let us assume the mean is constant, i.e.,
%
\begin{align}
    \pmb{\mu} = \mu \, \mathbf{j}
\end{align}
%
where $\mathbf{j}$ is the vector of $K$ ones.
%
Now suppose we cannot observe samples of $\mathbf{x}$ directly, but instead we can
observe samples from the \emph{normalized} process, which we will call
$\tilde{\mathbf{x}}$:
%
\begin{align}
    \tilde{\mathbf{x}}
     & \equiv \frac{\mathbf{x}}{\bar{x}}
    \nonumber                                                    \\[0.5em]
     & = \frac{\mathbf{x}}{\frac{1}{K}\sum\limits_{k=0}^{K-1} x}
    \quad.
\end{align}
%
The random variable $\tilde{\mathbf{x}}$ is no longer normally distributed, but as long as
the variance of $\bar{x}$ is small, we can attempt to approximate it as such. Let the mean
and covariance of $\tilde{\mathbf{x}}$ be $\tilde{\pmb{\mu}}$ and $\tilde{\pmb{\Sigma}}$, respectively:
%
\begin{align}
    \tilde{\mathbf{x}} \mathrel{\dot\sim} \mathcal{N}\left( \tilde{\pmb{\mu}}, \tilde{\pmb{\Sigma}} \right)
    \quad.
\end{align}
%
By definition, the mean is unity, i.e,
%
\begin{align}
    \tilde{\pmb{\mu}} = \mathbf{j}
    \quad.
\end{align}
%
What is the expression for $\tilde{\pmb{\Sigma}}$?

@@

    \tilde{\pmb{\Sigma}}

@@

    \mathbb{E}\big[ (\tilde{\mathbf{x}} - \tilde{\pmb{\mu}}) (\tilde{\mathbf{x}} - \tilde{\pmb{\mu}} )^\top \big]

@@

    \mathbb{E}\left[
        \left(\frac{\mathbf{x}}{\bar{x}} - \mathbf{j}\right)
        \left(\frac{\mathbf{x}}{\bar{x}} - \mathbf{j}\right)^\top

@@

    \mathbb{E}\left[
        \frac{\mathbf{x}\mathbf{x}^\top}{\bar{x}^2}

@@

        \frac{\mathbf{x}\,\mathbf{j}^\top}{\bar{x}}

@@

        \frac{\mathbf{j}\,\mathbf{x}^\top}{\bar{x}}

@@

    \mathbb{E}\left[
        \frac{\mathbf{x}\,\mathbf{j}^\top}{\bar{x}}

@@

    \mathbb{E}\left[
        \frac{\mathbf{j}\,\mathbf{x}^\top}{\bar{x}}

@@

    \mathbf{x} = \pmb{\mu} + \mathbf{L} \mathbf{u}

@@

and $\mathbf{u}$ is a standard multivariate normal random variable,

@@

    \mathbf{u} \sim \mathcal{N}\left( 0, \mathbf{I} \right)

@@

The expression for the covariance $\tilde{\pmb{\Sigma}}$ may now be written

@@

    \mathbb{E}\left[

@@

            (\pmb{\mu} + \mathbf{L} \mathbf{u})(\pmb{\mu} + \mathbf{L} \mathbf{u})^\top

@@

            \bar{x}^2

@@

            (\pmb{\mu} + \mathbf{L} \mathbf{u})\mathbf{j}^\top

@@

            \bar{x}

@@

            \mathbf{j}(\pmb{\mu} + \mathbf{L} \mathbf{u})^\top

@@

The mean $\bar{x}$ is

@@

    \bar{x} & = \sum\limits_{j=0}^{K-1}(\mu_j + L_{i,j}u_j)

@@

    \mu + \frac{1}{K}\mathbf{j}^\top \mathbf{L} \, \mathbf{u}

@@

            & = \mu(1 + \epsilon)

@@

    \epsilon
     & \equiv \frac{\bar{x}}{\mu} - 1

@@

     & = \frac{1}{\mu K}\mathbf{j}^\top \mathbf{L} \, \mathbf{u}

@@

    \mathbb{E}\Big[
        \frac{\epsilon^2 \, \mathbf{j} \, \mathbf{j}^\top}{(1 + \epsilon)^2}

@@

    \mathbb{E}\left[
        \frac{\epsilon\, \mathbf{L} \, \mathbf{u} \, \mathbf{j}^\top}{(1 + \epsilon)^2}

@@

    \mathbb{E}\left[
        \frac{ \mathbf{L} \, \mathbf{u} \, \mathbf{u}^\top \, \mathbf{L}^\top}{(1 + \epsilon)^2}

@@

$\big|\epsilon| < 1$ (an assumption we'll revisit below), we can Taylor expand the matrices as

@@

    \mathbb{E}\left[
        \epsilon^{n + 2}

@@

    \mathbb{E}\left[
        \epsilon^{n + 1} \,

@@

        \mathbf{u}

@@

    \mathbb{E}\left[
        \epsilon^n \, \mathbf{L} \, \mathbf{u} \, \mathbf{u}^\top \mathbf{L}^\top

@@

    \mathbb{E}\left[
        \epsilon^{n + 2} \, \mathbf{j} \, \mathbf{j}^\top

@@

    \mathbb{E}\left[
        \epsilon^{n + 1} \, \mathbf{L} \, \mathbf{u} \, \mathbf{j}^\top

@@

Why is this? Recall the assumption we made when expanding the expectation
integrals in Equations~(\ref{eq:WhereItAllGoesWrong1}), (\ref{eq:WhereItAllGoesWrong2}),
and (\ref{eq:WhereItAllGoesWrong3}): the Taylor expansion converges only for
$\big|\epsilon| < 1$. The quantity $\epsilon$ (Equation~\ref{eq:epsilon})
is indeed small provided the sample mean $\bar{x}$ is close to the GP
mean $\mu$. However, $\epsilon$ is a random variable, equal to
the weighted sum of $K$ standard normal random variables. Because a normal
distribution has infinite support, $\epsilon$ is guaranteed to take on
values greater than unity in the limit that an infinite number of samples
are drawn from the process. In particular, it is guaranteed to take on values
arbitrarily close to unity, in which case the sample mean approaches zero and
the values in the normalized sample diverge.
This reveals a fundamental flaw in our premise:
it is simply not correct to normalize a sample from a Gaussian process by its
mean value, since the resulting covariance is formally infinite.

@@

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/periodic_error.pdf}
        \oscaption{periodic_error}{%
            The covariance of the same normalized process as in
            Figure~\ref{fig:periodic}, but computed numerically
            from $10^5$ samples from the GP
            (left). The right panel shows the relative difference
            between the series solution and the numerical solution,
            normalized to the largest element in $\pmb{\tilde{\Sigma}}$.
            The maximum difference between the two solutions is less
            than one percent and is within the error of the numerical
            estimate.
            \label{fig:periodic_error}
        }
    \end{centering}
\end{figure}

>>>>

In this note, we show that these two common practices are somewhat at odds
with each other. Specifically, if a physical process that generates
a dataset is distributed as a GP, the normalized process \emph{cannot} be
distributed as a GP. Provided certain conditions are met, the normalized
process can be well \emph{approximated} by a GP, albeit one with a different
covariance matrix $\btilde{\pmb{\Sigma}}$ that is not simply a scalar
multiple of the original covariance matrix. Moreover, if the original process
is described by a stationary kernel (i.e., one in which covariances are
independent of phase), the normalized process is not guaranteed to be.

@@

Let $\mathbbb{x} = \left( \mathbb{x}_0 \,\, \cdots \,\, \mathbb{x}_{K-1} \right)^\top$
be a $K$-dimensional multivariate normal random variable distributed
according to a Gaussian Process with mean $\pmb{\mu}$
and covariance $\pmb{\Sigma}$:
%
\begin{align}
    \mathbbb{x} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad.
\end{align}
%
For simplicity, let us assume the mean is constant, i.e.,
%
\begin{align}
    \pmb{\mu} = \mu \, \mathbf{j}
\end{align}
%
where $\mathbf{j}$ is the vector of $K$ ones.
%
Now suppose we cannot observe samples of $\mathbbb{x}$ directly, but instead we can
observe samples from the \emph{normalized} process, which we will call
$\btilde{\mathbbb{x}}$:
%
\begin{align}
    \btilde{\mathbbb{x}}
     & \equiv \frac{\mathbbb{x}}{\bbar{\mathbb{x}}}
    \nonumber                                                              \\[0.5em]
     & = \frac{\mathbbb{x}}{\frac{1}{K}\sum\limits_{k=0}^{K-1} \mathbb{x}}
    \quad.
\end{align}
%
The random variable $\btilde{\mathbbb{x}}$ is no longer normally distributed, but as long as
the variance of $\bbar{\mathbb{x}}$ is small, we can attempt to approximate it as such. Let the mean
and covariance of $\btilde{\mathbbb{x}}$ be $\btilde{\pmb{\mu}}$ and $\btilde{\pmb{\Sigma}}$, respectively:
%
\begin{align}
    \btilde{\mathbbb{x}} \mathrel{\dot\sim} \mathcal{N}\left( \btilde{\pmb{\mu}}, \btilde{\pmb{\Sigma}} \right)
    \quad.
\end{align}
%
By definition, the mean is unity, i.e,
%
\begin{align}
    \btilde{\pmb{\mu}} = \mathbf{j}
    \quad.
\end{align}
%
What is the expression for $\btilde{\pmb{\Sigma}}$?

@@

    \btilde{\pmb{\Sigma}}

@@

    \mathbbb{E}\big[ (\btilde{\mathbbb{x}} - \btilde{\pmb{\mu}}) (\btilde{\mathbbb{x}} - \btilde{\pmb{\mu}} )^\top \big]

@@

    \mathbbb{E}\left[
        \left(\frac{\mathbbb{x}}{\bbar{\mathbb{x}}} - \mathbf{j}\right)
        \left(\frac{\mathbbb{x}}{\bbar{\mathbb{x}}} - \mathbf{j}\right)^\top

@@

    \mathbbb{E}\left[
        \frac{\mathbbb{x}\mathbbb{x}^\top}{\bbar{\mathbb{x}}^2}

@@

        \frac{\mathbbb{x}\,\mathbf{j}^\top}{\bbar{\mathbb{x}}}

@@

        \frac{\mathbf{j}\,\mathbbb{x}^\top}{\bbar{\mathbb{x}}}

@@

    \mathbbb{E}\left[
        \frac{\mathbbb{x}\,\mathbf{j}^\top}{\bbar{\mathbb{x}}}

@@

    \mathbbb{E}\left[
        \frac{\mathbf{j}\,\mathbbb{x}^\top}{\bbar{\mathbb{x}}}

@@

    \mathbbb{x} = \pmb{\mu} + \mathbf{L} \mathbbb{u}

@@

and $\mathbbb{u}$ is a standard multivariate normal random variable,

@@

    \mathbbb{u} \sim \mathcal{N}\left( 0, \mathbf{I} \right)

@@

The expression for the covariance $\btilde{\pmb{\Sigma}}$ may now be written

@@

    \mathbbb{E}\left[

@@

            (\pmb{\mu} + \mathbf{L} \mathbbb{u})(\pmb{\mu} + \mathbf{L} \mathbbb{u})^\top

@@

            \bbar{\mathbb{x}}^2

@@

            (\pmb{\mu} + \mathbf{L} \mathbbb{u})\mathbf{j}^\top

@@

            \bbar{\mathbb{x}}

@@

            \mathbf{j}(\pmb{\mu} + \mathbf{L} \mathbbb{u})^\top

@@

The mean $\bbar{\mathbb{x}}$ is

@@

    \bbar{\mathbb{x}} & = \sum\limits_{j=0}^{K-1}(\mu_j + L_{ij} \mathbb{u}_j)

@@

    \mu + \frac{1}{K}\mathbf{j}^\top \mathbf{L} \, \mathbbb{u}

@@

                      & = \mu(1 + \mathbb{r})

@@

    \mathbb{r}
     & \equiv \frac{\bbar{\mathbb{x}}}{\mu} - 1

@@

     & = \frac{1}{\mu K}\mathbf{j}^\top \mathbf{L} \, \mathbbb{u}

@@

    \mathbbb{E}\Big[
        \frac{\mathbb{r}^2 \, \mathbf{j} \, \mathbf{j}^\top}{(1 + \mathbb{r})^2}

@@

    \mathbbb{E}\left[
        \frac{\mathbb{r}\, \mathbf{L} \, \mathbbb{u} \, \mathbf{j}^\top}{(1 + \mathbb{r})^2}

@@

    \mathbbb{E}\left[
        \frac{ \mathbf{L} \, \mathbbb{u} \, \mathbbb{u}^\top \, \mathbf{L}^\top}{(1 + \mathbb{r})^2}

@@

$\big|\mathbb{r}| < 1$ (an assumption we'll revisit below), we can Taylor expand the matrices as

@@

    \mathbbb{E}\left[
        \mathbb{r}^{n + 2}

@@

    \mathbbb{E}\left[
        \mathbb{r}^{n + 1} \,

@@

        \mathbbb{u}

@@

    \mathbbb{E}\left[
        \mathbb{r}^n \, \mathbf{L} \, \mathbbb{u} \, \mathbbb{u}^\top \mathbf{L}^\top

@@

    \mathbbb{E}\left[
        \mathbb{r}^{n + 2} \, \mathbf{j} \, \mathbf{j}^\top

@@

    \mathbbb{E}\left[
        \mathbb{r}^{n + 1} \, \mathbf{L} \, \mathbbb{u} \, \mathbf{j}^\top

@@

Why is this? Recall the assumption we made when expanding the expectation
integrals in Equations~(\ref{eq:WhereItAllGoesWrong1}), (\ref{eq:WhereItAllGoesWrong2}),
and (\ref{eq:WhereItAllGoesWrong3}): the Taylor expansion converges only for
$\big|\mathbb{r}| < 1$. The quantity $\mathbb{r}$ (Equation~\ref{eq:epsilon})
is indeed small provided the sample mean $\bbar{\mathbb{x}}$ is close to the GP
mean $\mu$. However, $\mathbb{r}$ is a random variable, equal to
the weighted sum of $K$ standard normal random variables. Because a normal
distribution has infinite support, $\mathbb{r}$ is guaranteed to take on
values greater than unity in the limit that an infinite number of samples
are drawn from the process. In particular, it is guaranteed to take on values
arbitrarily close to unity, in which case the sample mean approaches zero and
the values in the normalized sample diverge.
This reveals a fundamental flaw in our premise:
it is simply not correct to normalize a sample from a Gaussian process by its
mean value, since the resulting covariance is formally infinite.

@@

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/periodic_error.pdf}
        \oscaption{periodic_error}{%
            The covariance of the same normalized process as in
            Figure~\ref{fig:periodic}, but computed numerically
            from $10^5$ samples from the GP
            (left). The right panel shows the relative difference
            between the series solution and the numerical solution,
            normalized to the largest element in $\btilde{\pmb{\Sigma}}$.
            The maximum difference between the two solutions is less
            than one percent and is within the error of the numerical
            estimate.
            \label{fig:periodic_error}
        }
    \end{centering}
\end{figure}
